# **Week 6 - Wrap up skills portion (11/06/2023)**

Browse()

Browse *pauses execution*, allowing you to look around and explore.

## **Error in `summarize()`**

Andrea needs to update her RStudio. The errors that the next exercise should show do not show correctly.

The following pipeline has an error. Use it to answer the following questions.

```{r}
library(palmerpenguins)
library(tidyverse)

traits_r2 <- function(trait1, trait2) {
  summary(lm(trait1 ~ trait2))$rsquared
}

penguins %>% 
  summarize(flipper_bill_r2 = traits_r2(flipper_length_mm, bill_length_mm))
```

Q1: How would you describe the error? I\'m not asking you describe the *cause* of the error yet. Describe how the output is different than what you would expect. Expected rows. There is not data in the tibble

**Q2:** Where would you add `browser()` to explore the cause of the error? In the function "traits_r2".

**Q3:** Does the body of `traits_r2` use list or atomic vector indexing? Does it use indexing by position, logic, or name? Indexing by name.

**Q4:** What\'s the cause of the error? How would you fix it? rsquared needed a period = "r.squared"

## **Adding `group_by()`**

The following pipeline is similar to the one above, with an added layer of complexity. Use it to answer the following questions.

```{r}
# Pipeline 1
penguins %>% 
  group_by(species) %>% 
  summarize(flipper_bill_r2 = traits_r2(flipper_length_mm, bill_length_mm))

# Pipeline 2
penguins %>% 
  group_by(species, island) %>% 
  summarize(flipper_bill_r2 = traits_r2(flipper_length_mm, bill_length_mm),
            .groups = "drop")
```

**Q5:** How many times does Pipeline 1 call `traits_r2()`? Output tibble shows that it was called 3 times, one per each species. How about Pipeline 2? 5 times.

**Q6:** Create Pipeline 3 that additionally groups by sex. How many times does Pipeline 3 call `traits_r2()`? 13 times.

```{r}
#Pipeline 3
penguins %>% 
  group_by(species, island,sex) %>% 
  summarize(flipper_bill_r2 = traits_r2(flipper_length_mm, bill_length_mm),
            .groups = "drop")
```

## **Error in `group_by()`-`summarize()`**

The following code creates an error in Pipeline 3. Change your Pipeline 3 to use `penguins2` instead of `penguins`, then answer the following questions.

```{r}
traits_r2 <- function(trait1, trait2) {
  #browse()
  summary(lm(trait1 ~ trait2))$r.squared #fixed
}
```

```{r}
set.seed(12345) #starts random numer generation at a fixed point.
penguins2 <- penguins %>% 
  drop_na(sex) %>% 
  sample_n(25)
penguins2[7, 3:6] <- NA

#Pipeline 3
penguins2 %>% 
  group_by(species, island,sex) %>% 
  drop_na()%>%
  summarize(flipper_bill_r2 = traits_r2(flipper_length_mm, bill_length_mm),
            .groups = "drop")

```

**Q7:** How would you describe the error? NA in group 6 (in Trait1 and Trait 2) when running browse() is right after the bracket in the function "traits_r2"

**Q8:** Use `browser()` to diagnose the error. Hint: `c` will tell the debugger to continue until the next time it\'s called.

**Q9:** How would you fix the error? in penguins2 add drop\_ na() after the first pipeline, or before summarize. Final output = 7 groups.

# **Visualizing data with `ggplot()`**

This is a basic introduction to data visualization with `ggplot()`, which should come in handy when you\'re creating figures for your final projects.

## **Components of `ggplot()`**

`ggplot()` has four major components:

-   Data

-   Scales

-   Geometries

-   Aesthetics

### **Data**

A **data frame** containing the data you want to visualize. Emphasis on **data frame**. `ggplot()` *is not designed to work with individual vectors*, which you might be used to if you\'ve used base R, Matlab, or other plotting tools.

### **Scales**

Scales determine **where/how** variables in your data will show up in the plot. These include the x- and y-axes, as well as color, size, and more.

### **Geometries**

Geometries determine the **geometric properties** of your variables. E.g. scatter plots use a point geometry and line plots use a line geometry. Other familiar geometries include histograms and box plots.

### **Aesthetics**

The aesthetics unify data, scales, and geometries. They tell `ggplot()` how to translate data to scales for each geometry.

## **A simple example**

```{r}
library(palmerpenguins)
library(tidyverse)

ggplot(data = penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point()

head(penguins)
```

**Q10:** Change the aesthetics of the plot to show `bill_depth_mm` on the y-axis.

```{r}
ggplot(data = penguins, aes(x = body_mass_g, y = bill_depth_mm)) +
  geom_point()
```

**Q11:** Add an aesthetic for color, so points will be color-coded by species.

```{r}
ggplot(data = penguins, aes(x = body_mass_g, y = bill_depth_mm)) +
  geom_point(aes(col = species))
```

## **Customizing scales**

In addition to adding geometries with `geom_*()`, we can add scales with `scale_*_*()` to customize how scales appear in the figure. The first `*` is the name of the aesthetic, and the second `*` is the type of scale.

```{r}
ggplot(data = penguins) +
  # You can also define aesthetics directly inside the geometries
  geom_point(aes(x = body_mass_g, 
                 y = flipper_length_mm,
                 color = species)) +
  # x aesthetic, continuous scale
  scale_x_continuous(
    # change the axis name
    name = "Body mass (g)",
    # change the limits
    limits = c(2000, 8000)
  ) +
  # color aesthetic, manual scale
  scale_color_manual(
    # set the values for the colors
    values = c(Adelie = "cornflowerblue",
               Chinstrap = "firebrick",
               Gentoo = "darkorchid")
  )
```

**Q12:** What function would you use to customize the y-axis? Use that function to expand the y-axis limits to include 150 and 250.

```{r}
ggplot(data = penguins) +
  # You can also define aesthetics directly inside the geometries
  geom_point(aes(x = body_mass_g, 
                 y = flipper_length_mm,
                 color = species)) +
  # x aesthetic, continuous scale
  scale_x_continuous(
    # change the axis name
    name = "Body mass (g)",
    # change the limits
    limits = c(2000, 8000)
  ) +
  # y aesthetic, continuous scale
  scale_y_continuous(
    # change the axis name
    name = "Body mass (g)",
    # change the limits
    limits = c(150, 250)
  ) +
  # color aesthetic, manual scale
  scale_color_manual(
    # set the values for the colors
    values = c(Adelie = "cornflowerblue",
               Chinstrap = "firebrick",
               Gentoo = "darkorchid")
  )
```

**Q13:** Look up the help for `scale_color_brewer()`. Change the color scale to use the \"Dark2\" color palette.

```{r}
ggplot(data = penguins) +
  # You can also define aesthetics directly inside the geometries
  geom_point(aes(x = body_mass_g, 
                 y = flipper_length_mm,
                 color = species)) +
  # x aesthetic, continuous scale
  scale_x_continuous(
    # change the axis name
    name = "Body mass (g)",
    # change the limits
    limits = c(2000, 8000)
  ) +
  # y aesthetic, continuous scale
  scale_y_continuous(
    # change the axis name
    name = "Body mass (g)",
    # change the limits
    limits = c(150, 250)
  ) +
  # color aesthetic using "scale_colour_brewer" pallete "Dark2"
  scale_colour_brewer(palette = "Dark2")
  )
```

## **Other useful geometries**

**Q14:** Use `geom_histogram()` to visualize the distribution of body masses.

```{r}
ggplot(data=penguins) +
  geom_histogram(aes(body_mass_g))
```

**Q15:** Use `geom_boxplot()` to create boxplots of the flipper lengths by species. It should look like the figure below. Hint: what aesthetics will you use?

```{r}
ggplot(data=penguins, aes(x=species, y=flipper_length_mm))+
  geom_boxplot()
```

**Q16:** `geom_pointrange()` is like `geom_point()` with the addition of error bars. Like `geom_point()`, `geom_pointrange()` takes `x` and `y` aesthetics, but it also takes `xmin`, `xmax`, `ymin` and `ymax` aesthetics. Create a figure that shows the mean ± sd of each penguin species\' body mass and flipper lengths using points and error bars; it should look like the figure below. First, create a summary data frame that calculates the mean and standard deviation of the two variables you need by each species. Then use that summary data frame to create your figure. Hint: `geom_pointrange()` takes `xmin` ***or*** `xmax`, same for `ymin` ***or*** `ymax`. So you\'ll need to call `geom_pointrange()` twice and specify the aesthetics within the geometry.

```{r}

#summarize data
penguins %>% 
  group_by(species) %>% 
  drop_na() %>%
    summarize(mean_m = mean(body_mass_g),
              sd_m = sd(body_mass_g),
              mean_f = mean(flipper_length_mm),
              sd_f = sd(flipper_length_mm)) %>%
#plot with xmax xmin, ymax, ymin, and SDs
ggplot(aes(color = species))+
  geom_pointrange(aes(x=mean_m,
                      y=mean_f,
                      xmin=mean_m-sd_m,
                      xmax=mean_m+sd_m))+
  geom_pointrange(aes(x=mean_m,
                      y=mean_f,
                      ymin=mean_f-sd_f,
                      ymax=mean_f+sd_f))
```
